---
title: "afsc_combo_survey"
author: "Joshua Zahner"
date: "2/12/2023"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# NWFSC Slope Survey

```{r}
library(nwfscSurvey)
library(tidyverse)

species <- "shortspine thornyhead"
survey.name <- "AFSC.Slope"
survey.name.path <- str_replace(tolower(survey.name), "[.]", "_")

data.dir <- here::here("data/") # location of data directory
outputs.dir <- here::here("outputs/")

raw.data.dir <- file.path(data.dir, "raw")
catch.fname <- file.path(raw.data.dir, paste0(survey.name.path, "_survey_catch.csv")) # raw survey catch filename
bio.lengths.fname <- file.path(raw.data.dir, paste0(survey.name.path, "_survey_bio_lengths.csv"))     # raw survey bio filename
bio.ages.fname <- file.path(raw.data.dir, paste0(survey.name.path, "_survey_bio_ages.csv"))     # raw survey bio filename
out.dir <- file.path(outputs.dir, "surveys", survey.name.path) # send all plot/data outputs here
```

```{r}
dir.create(raw.data.dir, recursive=TRUE)
dir.create(out.dir, recursive=TRUE)
```


```{r, echo=FALSE, message=FALSE, results="hide"}
# Fancy way of automatically using the raw data if its available as a file
# or pulling it from the server and saving it if its not already there.

if(!file.exists(catch.fname) | !file.exists(bio.fname)){
    catch <- PullCatch.fn(Name = species, 
                     SurveyName = survey.name)
                     
    bio   <- PullBio.fn(Name = species, 
                       SurveyName = survey.name)
    
    write.csv(catch, catch.fname)
    write.csv(bio$Lengths, bio.lengths.fname)
    write.csv(bio$Ages, bio.ages.fname )
}

catch <- read.csv(catch.fname)
bio$Lengths <- read.csv(bio.lengths.fname)
bio$Ages <- read.csv(bio.ages.fname)

head(catch)
head(bio)
```

```{r}
plot_cpue(
  dir = out.dir, 
  catch = catch
)

plot_bio_patterns(
  dir = out.dir, 
  bio = bio$Lengths, 
  col_name = "Length_cm")

PlotMap.fn(dir=out.dir, dat=catch)

PlotSexRatio.fn(dir=out.dir, bio$Lengths, data.type="length")
PlotSexRatio.fn(dir=out.dir, bio$Ages, data.type="age")

```
CPUE appears relatively flat across latitude for which data is available. Extreme differences in latitude included in the survey across years is obvious, with the survey concentrated in mid to high latitudes until 1997, after which it covered the whole of the west coast. There does not appear to be any major trends in CPUE by depth. *Going to exclude pre-1997 data for now, but this should be revisited.*

Lots of available length data from 1988 onward, with obvious trend of larger individuals in deeper waters. Some unsexed, but mostly all small.

```{r}
catch <- catch %>% filter(Year >= 1997)
bio$Lengths <- bio$Lengths %>% filter(Year >= 1997)
bio$Ages <- bio$Ages %>% filter(Year > 1997)
```


```{r}
# Made a small change to the default wh_plot_proportion() function in nwfscSurvey so
# that the plots are shown in line as well as saved to files.
wh_plot_proportion <- function (data_catch, data_bio, dir = file.path(getwd(), "plots"), 
    bar_width = c("n", "equal")) 
{
    stopifnot(any(c(!missing(data_catch), !missing(data_bio))))
    files_all <- file.path(dir, t(outer(X = c(if (!missing(data_catch)) {
        "presence-absence"
    }, if (!missing(data_bio)) {
        "sex"
    }), Y = paste0("_by_", c("depth", "latitude"), ".png"), FUN = paste0)))
    data <- c(if (!missing(data_catch)) {
        dplyr::mutate(data_catch, the_factor = factor(cpue_kg_km2 <= 
            0, levels = c(FALSE, TRUE), labels = c("Present", 
            "Absent"))) %>% purrr::rerun(.n = 2)
    }, if (!missing(data_bio)) {
        dplyr::mutate(data_bio, the_factor = codify_sex(Sex)) %>% 
            purrr::rerun(.n = 2)
    })
    gg_all <- purrr::pmap(.l = list(pdata = data, x = rep(ggplot2::quos(Depth_m, 
        Latitude_dd), length(data)/2), width = rep(c(50, 1), 
        length(data)/2)), .f = function(pdata, x, width, bar_width) {
        gg <- plot_proportion(data = pdata, column_factor = the_factor, 
            column_bin = !!x, width = width, bar_width = bar_width, 
            boundary = 0)
    }, bar_width = match.arg(bar_width))
    files_out <- purrr::map2_chr(.x = files_all, .y = gg_all, 
        .f = ggplot2::ggsave, height = 7, width = 7)
    return(list(files_out, gg_all))
}

wh_plot_proportion(
  dir = file.path(out.dir, "plots"),
  data_catch = catch,
  data_bio = bio$Lengths
)
```
No obvious presence-absence or sex-ratio trends across latitude or depth present.

```{r}
# Set spatial strata
strata = CreateStrataDF.fn(
  names          = c("shallow_south", "deep_south"),
  depths.shallow = c(55,    500),
  depths.deep    = c(500,  1280),
  lats.south     = c(34,     34),
  lats.north     = c(49, 49)
)

```
Strata are set to the same as they were in 2014 (p. 27).
I dont see any particularly strong reason to change the latitudinal bins, since CPUE and presence-absence are very consistent over latitude.
Only using two depth bins seems like it could be revised though. Maybe use 55-350 m, 350-750 m, 750+ m?


## Design Based Indices
```{r}
## Calculate design-based index

biomass.idx = Biomass.fn(dir = out.dir,
                         dat = catch,
                         strat.df = strata)

PlotBio.fn(
  dir = out.dir, # set to `out.dir`to save as file
  dat = biomass.idx,
  scalar = 1000, # convert to kg like csv file
) # produces '_designed_based_index.png'

PlotBioStrata.fn(
  dir = out.dir, # set to `out.dir`to save as file
  dat = biomass.idx,
  scalar = 1000, # convert to kg like csv file
) # produces '_designed_based_by_strata_index.png'

```


## Length Compositions

```{r}
n.len.raw <- nrow(bio$Lengths[!is.na(bio$Lengths$Length_cm),]) # check if data available

if(n.len.raw > 0){
    n <- GetN.fn(dir = out.dir, 
         dat = bio$Lengths, 
         type = "length", 
         species = "thorny")
      
    len_bins <- seq(6, 72, 2) # from last assessment (p. 125)
    
    Length_Freq <- SurveyLFs.fn(dir = out.dir, 
                            datL =  bio$Lengths, 
                            datTows = catch,
                            strat.df = strata,
                            lgthBins = len_bins)
    
    
    PlotFreqData.fn(dir = out.dir, # set to `out.dir`to save as file 
                    dat = Length_Freq)
    
    plot_comps(dir = out.dir, data = Length_Freq) # uses version in R/utils/ 
}else{
  print(n.len.raw)
}

```
Lengths look generally consistent with other surveys. Effective sample sizes are all high (>1000).


## Age Compositions

```{r}
n.age.raw <- nrow(bio$Ages[!is.na(bio$Ages$Age),]) # check is data available

if(n.age.raw > 0){
  
  n <- GetN.fn(dir = out.dir,
         dat = bio$Ages,
         type = "age",
         species = "thorny")

  age_bins <- 1:70 # arbitrarily set right now as not used in previous assessment
  
  Ages <- SurveyAFs.fn(dir = out.dir,
                       datA = bio$Ages,
                       datTows = catch,
                       strat.df = strata,
                       ageBins = age_bins,
                       nSamps = n)
  
  PlotFreqData.fn(
    dir = out.dir,
    dat = Ages
  )
  
  plot_comps(
    dir = out.dir,
    data = Ages
  ) # uses the version in R/utils/
  
}else{
  print(n.age.raw)
}

```
Age data is available prior to 1997, but the composition seems suspect at best, regardless of the high effective sample sizes. Probably shouldn't use.

