---
title: "afsc_triennial1_survey"
author: "Joshua Zahner"
date: "2/03/2023"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# AFSC Triennial 1 Survey

```{r}
library(nwfscSurvey)
library(tidyverse)

species <- "shortspine thornyhead"
survey.name <- "Triennial"
survey.name.path <- str_replace(tolower(survey.name), "[.]", "_")
survey.name.path <- paste0(survey.name.path, "1")

data.dir <- here::here("data/") # location of data directory
outputs.dir <- here::here("outputs/")

raw.data.dir <- file.path(data.dir, "raw")
catch.fname <- file.path(raw.data.dir, paste0(survey.name.path, "_survey_catch.csv")) # raw survey catch filename
bio.fname <- file.path(raw.data.dir, paste0(survey.name.path, "_survey_bio.csv"))     # raw survey bio filename
out.dir <- file.path(outputs.dir, "surveys", survey.name.path) # send all plot/data outputs here
```

```{r}
dir.create(raw.data.dir, recursive=TRUE)
dir.create(out.dir, recursive=TRUE)
```


```{r, echo=FALSE, message=FALSE, results="hide"}
# Fancy way of automatically using the raw data if its available as a file
# or pulling it from the server and saving it if its not already there.

if(!file.exists(catch.fname) | !file.exists(bio.fname)){
    catch <- PullCatch.fn(Name = species, 
                     SurveyName = survey.name)
                     
    bio   <- PullBio.fn(Name = species, 
                       SurveyName = survey.name)
    
    catch <- catch %>% filter(Depth_m <= 366)               # This is the same assumption made in 2013
    bio$Lengths <- bio$Lengths %>% filter(Depth_m <= 366)   # This is the same assumption made in 2013

    
    write.csv(catch, catch.fname)
    write.csv(bio$Lengths, bio.fname)
}

catch <- read.csv(catch.fname)
bio$Lengths <- read.csv(bio.fname)

head(catch)
head(bio$Lengths)
```

```{r}
plot_cpue(
  dir = out.dir, 
  catch = catch
)

plot_bio_patterns(
  dir = out.dir, 
  bio = bio$Lengths, 
  col_name = "Length_cm")

PlotMap.fn(dir=out.dir, dat=catch)

PlotSexRatio.fn(dir=out.dir, bio$Lengths, data.type="length")
#PlotSexRatio.fn(dir=out.dir, bio, data.type="age") # No ages available

```

No clear CPUE trend over time across Latitude, though CPUE map clearly shows that effort in early years was concentrated in Washington an Oregon, and effort only equalized into California in later years. There is a clear positive correlation between CPUE and Depth, with larger depths consistently receiving additional effort. 

Plenty of sexed data is available from 1995 -2004, though many small unsexed fish. Few sexed samples in 1980, 1983, and 1986 to work with, and spotty sexing coverage in 1992.

```{r}
# Made a small change to the default wh_plot_proportion() function in nwfscSurvey so
# that the plots are shown in line as well as saved to files.
wh_plot_proportion <- function (data_catch, data_bio, dir = file.path(getwd(), "plots"), 
    bar_width = c("n", "equal")) 
{
    stopifnot(any(c(!missing(data_catch), !missing(data_bio))))
    files_all <- file.path(dir, t(outer(X = c(if (!missing(data_catch)) {
        "presence-absence"
    }, if (!missing(data_bio)) {
        "sex"
    }), Y = paste0("_by_", c("depth", "latitude"), ".png"), FUN = paste0)))
    data <- c(if (!missing(data_catch)) {
        dplyr::mutate(data_catch, the_factor = factor(cpue_kg_km2 <= 
            0, levels = c(FALSE, TRUE), labels = c("Present", 
            "Absent"))) %>% purrr::rerun(.n = 2)
    }, if (!missing(data_bio)) {
        dplyr::mutate(data_bio, the_factor = codify_sex(Sex)) %>% 
            purrr::rerun(.n = 2)
    })
    gg_all <- purrr::pmap(.l = list(pdata = data, x = rep(ggplot2::quos(Depth_m, 
        Latitude_dd), length(data)/2), width = rep(c(50, 1), 
        length(data)/2)), .f = function(pdata, x, width, bar_width) {
        gg <- plot_proportion(data = pdata, column_factor = the_factor, 
            column_bin = !!x, width = width, bar_width = bar_width, 
            boundary = 0)
    }, bar_width = match.arg(bar_width))
    files_out <- purrr::map2_chr(.x = files_all, .y = gg_all, 
        .f = ggplot2::ggsave, height = 7, width = 7)
    return(list(files_out, gg_all))
}

wh_plot_proportion(
  dir = file.path(out.dir, "plots"),
  data_catch = catch,
  data_bio = bio$Lengths
)
```
Definitely a clear depth trend in presence-absence data with higher presence in deeper waters. Relatively consistent presence-absence across latitude.

Latitudinal strata can be set at convenience (state boundaries), but depth strata should be set carefully (probably a shallow and deep). Very surprised that the 2013 assessment modeled this as a single strata given how strong the presence-absence trend with depth is.

```{r}
# Set spatial strata
strata = CreateStrataDF.fn(
  depths.shallow = 100,
  depths.deep    = 400,
  lats.south     = 34.5,
  lats.north     = 49
)
```

## Design Based Indices
```{r}
## Calculate design-based index

biomass.idx = Biomass.fn(dir = out.dir, 
                         dat = catch,  
                         strat.df = strata)

PlotBio.fn(
  dir = out.dir, # set to `out.dir`to save as file
  dat = biomass.idx,
  scalar = 1000, # convert to kg like csv file
) # produces '_designed_based_index.png'

PlotBioStrata.fn(
  dir = out.dir, # set to `out.dir`to save as file
  dat = biomass.idx,
  scalar = 1000, # convert to kg like csv file
) # produces '_designed_based_by_strata_index.png'

```

## Length Compositions

```{r}
n.len.raw <- nrow(bio$Lengths[!is.na(bio$Lengths$Length_cm),]) # check is data available

if(n.len.raw > 0){
    n <- GetN.fn(dir = out.dir, 
         dat = bio$Lengths, 
         type = "length", 
         species = "thorny")
      
    len_bins <- seq(6, 72, 2) # from last assessment (p. 125)
    
    Length_Freq <- SurveyLFs.fn(dir = out.dir, 
                            datL =  bio$Lengths, 
                            datTows = catch,
                            strat.df = strata,
                            lgthBins = len_bins)
    
    
    PlotFreqData.fn(dir = out.dir, # set to `out.dir`to save as file 
                    dat = Length_Freq)
    
    plot_comps(dir = out.dir, data = Length_Freq) # uses version in R/utils/
    
    print(n)
    
}else{
  print(n.len.raw)
}

```
Length compositions look consistent with other surveys. pre-1989 comps are a bit jagged, likely due to small sample size, and may not be useful. Decent effective sample sizes from 1989 onwards, though 1992 is a bit low.

## Age Compositions

```{r}
n.age.raw <- tryCatch({nrow(bio$Ages[!is.na(bio$Ages$Age),])}, error=function(e){0}) # check is data available

if(n.age.raw > 0){
  
  n <- GetN.fn(dir = out.dir,
         dat = bio$Ages,
         type = "age",
         species = "thorny")

  age_bins <- 1:40 # arbitrarily set right now as not used in previous assessment
  
  Ages <- SurveyAFs.fn(dir = out.dir,
                       datA = bio$Ages,
                       datTows = catch,
                       strat.df = strata,
                       ageBins = age_bins,
                       nSamps = n)
  
  PlotFreqData.fn(
    dir = out.dir,
    dat = Ages
  )
  
  plot_comps(
    dir = out.dir,
    data = Ages
  ) # uses the version in R/utils/
  
}else{
  print(n.age.raw)
}

```

## Conditional Age-at-Length
There are no ages associated with this species and survey, so CAAL analyses are not possible.
```{r}
if(n.age.raw > 0){
  caal <- SurveyAgeAtLen.fn(dir = out.dir,
                          datAL = bio,
                          datTows = catch,
                          strat.df = strata,
                          lgthBins = len_bins,
                          ageBins = age_bins) 
}
```

