---
title: "nwfsc_combo_survey"
author: "Joshua Zahner"
date: "1/31/2023"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# NWFSC Slope Survey

```{r}
library(nwfscSurvey)
library(tidyverse)

species <- "shortspine thornyhead"
survey.name <- "NWFSC.Slope"
survey.name.path <- str_replace(tolower(survey.name), "[.]", "_")

data.dir <- here::here("data/") # location of data directory
outputs.dir <- here::here("outputs/")

raw.data.dir <- file.path(data.dir, "raw")
catch.fname <- file.path(raw.data.dir, paste0(survey.name.path, "_survey_catch.csv")) # raw survey catch filename
bio.fname <- file.path(raw.data.dir, paste0(survey.name.path, "_survey_bio.csv"))     # raw survey bio filename
out.dir <- file.path(outputs.dir, "surveys", survey.name.path) # send all plot/data outputs here
```

```{r}
dir.create(raw.data.dir, recursive=TRUE)
dir.create(out.dir, recursive=TRUE)
```


```{r, echo=FALSE, message=FALSE, results="hide"}
# Fancy way of automatically using the raw data if its available as a file
# or pulling it from the server and saving it if its not already there.

if(!file.exists(catch.fname) | !file.exists(bio.fname)){
    catch <- PullCatch.fn(Name = species, 
                     SurveyName = survey.name)
                     
    bio   <- PullBio.fn(Name = species, 
                       SurveyName = survey.name)
    
    write.csv(catch, catch.fname)
    write.csv(bio, bio.fname)
}

catch <- read.csv(catch.fname)
bio <- read.csv(bio.fname)

head(catch)
head(bio)
```

```{r}
plot_cpue(
  dir = out.dir, 
  catch = catch
)

plot_bio_patterns(
  dir = out.dir, 
  bio = bio, 
  col_name = "Length_cm")

PlotMap.fn(dir=out.dir, dat=catch)

PlotSexRatio.fn(dir=out.dir, bio, data.type="length")
PlotSexRatio.fn(dir=out.dir, bio, data.type="age")

```
CPUE seems to be pretty flat across latitude over time, though there is definitely a decrease in middle latitudes (~38˚-42˚). Maybe a slight positive trend in CPUE with depth (higher CPUE as greater depth). Definitely something going on in 2002.

Clear positive trend in length with depth, and no obvious trends between length and latitude. Indicates we need to be thoughful as to where to set depth strata, but latitudinal strata can be set pretty much anywhere.

NOTE that almost all samples in this survey are unsexed. Will need to be VERY careful how we deal with length compositions form this survey if we assume sex-specific growth parameters.

```{r}
# Made a small change to the default wh_plot_proportion() function in nwfscSurvey so
# that the plots are shown in line as well as saved to files.
wh_plot_proportion <- function (data_catch, data_bio, dir = file.path(getwd(), "plots"), 
    bar_width = c("n", "equal")) 
{
    stopifnot(any(c(!missing(data_catch), !missing(data_bio))))
    files_all <- file.path(dir, t(outer(X = c(if (!missing(data_catch)) {
        "presence-absence"
    }, if (!missing(data_bio)) {
        "sex"
    }), Y = paste0("_by_", c("depth", "latitude"), ".png"), FUN = paste0)))
    data <- c(if (!missing(data_catch)) {
        dplyr::mutate(data_catch, the_factor = factor(cpue_kg_km2 <= 
            0, levels = c(FALSE, TRUE), labels = c("Present", 
            "Absent"))) %>% purrr::rerun(.n = 2)
    }, if (!missing(data_bio)) {
        dplyr::mutate(data_bio, the_factor = codify_sex(Sex)) %>% 
            purrr::rerun(.n = 2)
    })
    gg_all <- purrr::pmap(.l = list(pdata = data, x = rep(ggplot2::quos(Depth_m, 
        Latitude_dd), length(data)/2), width = rep(c(50, 1), 
        length(data)/2)), .f = function(pdata, x, width, bar_width) {
        gg <- plot_proportion(data = pdata, column_factor = the_factor, 
            column_bin = !!x, width = width, bar_width = bar_width, 
            boundary = 0)
    }, bar_width = match.arg(bar_width))
    files_out <- purrr::map2_chr(.x = files_all, .y = gg_all, 
        .f = ggplot2::ggsave, height = 7, width = 7)
    return(list(files_out, gg_all))
}

wh_plot_proportion(
  dir = file.path(out.dir, "plots"),
  data_catch = catch,
  data_bio = bio
)
```
Presence-absence plots indicate the species is most prevalent in ~350-1000m of water, and is much prevalent in shallower waters. There are no obvious presence-absence trends across latitude. These both again imply that depth strata should be set thoughtfully, but that latitudinal strata can be set at convenience.

No trends in sex-specific presence-absence across depth or latitude is discernible due to the large proportion of unsexed fish.

```{r}
# Set spatial strata
strata = CreateStrataDF.fn(
  names          = c("shallow_south", "deep_south", "shallow_cen", "deep_cen", "shallow_north", "deep_north"), 
  depths.shallow = c(55,    500,  55,   500,   55,  500),
  depths.deep    = c(500,  1280,  500,  1280, 500, 1280),
  lats.south     = c(32,     32, 40.5,  40.5,  43,   43),
  lats.north     = c(40.5, 40.5,   43 ,   43,  49, 	 49)
)
```
Strata are set to the same as they were in 2014 (p. 27).
I dont see any particularly strong reason to change the latitudinal bins, since CPUE and presence-absence are very consistent over latitude.
Only using two depth bins seems like it could be revised though. Maybe use 55-350 m, 350-750 m, 750+ m?


## Design Based Indices
```{r}
## Calculate design-based index

biomass.idx = Biomass.fn(dir = out.dir,
                         dat = catch,
                         strat.df = strata)




PlotBio.fn(
  dir = NULL, # set to `out.dir`to save as file
  dat = biomass.idx,
  scalar = 1000, # convert to kg like csv file
) # produces '_designed_based_index.png'

PlotBioStrata.fn(
  dir = NULL, # set to `out.dir`to save as file
  dat = biomass.idx,
  scalar = 1000, # convert to kg like csv file
) # produces '_designed_based_by_strata_index.png'

```


## Length Compositions

Lengths only available for the year 2000. Male length comps look generally consistent with other surveys, though female comps look less similar. Could be due to low sample sizes or lack of sexed data from this survey. Length comps from this survey may not be of sufficient quality to warrant use in the assessment. 
```{r}
n.len.raw <- nrow(bio[!is.na(bio$Length_cm),]) # check if data available

if(n.len.raw > 0){
    n <- GetN.fn(dir = out.dir, 
         dat = bio, 
         type = "length", 
         species = "thorny")
      
    len_bins <- seq(6, 72, 2) # from last assessment (p. 125)
    
    Length_Freq <- SurveyLFs.fn(dir = out.dir, 
                            datL =  bio, 
                            datTows = catch,
                            strat.df = strata,
                            lgthBins = len_bins)
    
    
    PlotFreqData.fn(dir = NULL, # set to `out.dir`to save as file 
                    dat = Length_Freq)
    
    plot_comps(dir = out.dir, data = Length_Freq) # uses version in R/utils/ 
}else{
  print(n.len.raw)
}

```

## Age Compositions

Small samples sizes available for ages in 2000. Quality of data is unknown, but is likely suffers from similar sample size and sexing issues as the length composition data.
```{r}
n.age.raw <- nrow(bio[!is.na(bio$Age),]) # check is data available

if(n.age.raw > 0){
  
  n <- GetN.fn(dir = out.dir,
         dat = bio,
         type = "age",
         species = "thorny")

  age_bins <- 1:40 # arbitrarily set right now as not used in previous assessment
  
  Ages <- SurveyAFs.fn(dir = out.dir,
                       datA = bio,
                       datTows = catch,
                       strat.df = strata,
                       ageBins = age_bins,
                       nSamps = n)
  
  PlotFreqData.fn(
    dir = out.dir,
    dat = Ages
  )
  
  plot_comps(
    dir = out.dir,
    data = Ages
  ) # uses the version in R/utils/
  
}else{
  print(n.age.raw)
}

```

## Conditional Age-at-Length
```{r}
caal <- SurveyAgeAtLen.fn(dir = out.dir, 
                          datAL = bio, 
                          datTows = catch,
                          strat.df = strata,
                          lgthBins = len_bins, 
                          ageBins = age_bins)

plot_age_length_sampling(dir=out.dir, caal)

```

