---
title: "Catch only projection for shortspine thornyhead"
author: "Kiva L. Oken" 
affiliation: 'NOAA Fisheries Northwest Fisheries Science Center, 2725 Montlake Blvd E, Seattle, WA 98112'
date: today
format: 
  pdf:
    template: catch_only_projection_template.tex
params:
  species: "shortspine thornyhead"
  assess_year: 2023
  current_year: 2025
  final_proj_year: 2036
  category: 2
  sigma: 1
  p_star: 0.45
  spawn_output_decimals: 2
  dir_original: "../model/Sensitivity_Anal/Base_Model/5.25_Decision_Table_045/1_23.base.dt_base_45/run"
  dir_new: "cop1"
  dir_new2: "cop2b"
  dir_new3: "cop3"
  spawn_output_units: "(billions of eggs)"
---


```{r}
#| label: parameters
#| echo: false
#| warning: false
#| message: false
species <- params$species
Species <- stringr::str_to_sentence(params$species)
assess_year <- params$assess_year
current_year <- params$current_year
final_proj_year <- params$final_proj_year
category <- params$category
sigma <- params$sigma
p_star <- params$p_star
spawn_output_decimals <- params$spawn_output_decimals

# original model
model_original <- r4ss::SS_output(
    dir = params$dir_original,
    verbose = FALSE,
    printstats = FALSE,
    hidewarn = TRUE
)

# catch-only projection
model <- r4ss::SS_output(
    dir = params$dir_new, # this is the directory where I ran the standard Catch-only projection model
    verbose = FALSE,
    printstats = FALSE,
    hidewarn = TRUE
)

# catch-only projection Alternative 1
model2 <- r4ss::SS_output(
    dir = params$dir_new2, 
    verbose = FALSE,
    printstats = FALSE,
    hidewarn = TRUE
)

# catch-only projection Alternative 2
model3 <- r4ss::SS_output(
    dir = params$dir_new3,
    verbose = FALSE,
    printstats = FALSE,
    hidewarn = TRUE
)

table_caption <- glue::glue("Original {assess_year} projection and new catch-only projection for OFLs (mt), buffer, ABCs (mt), actual & assumed catch (mt), ACLs (mt), spawning output {params$spawn_output_units}, and stock status (fraction of unfished spawning output) given the removals. The gray shading indicates values associated with the original {assess_year} assessment projection. In both the original model and the new projection, the removals for years beyond the values in the 'Actual & Assumed Removals' columns are the ACL values estimated from that model.")

table_caption2 <- glue::glue("Original {assess_year} projection and alternative catch-only projection in which the 2026 assumed catches are set to 67% of a new ACL based on actual and assumed catches through 2025 rather than fixed at the values calculated by the GMT based on the current harvest specifications. See caption of table above for additional details.")

table_caption3 <- glue::glue("Original {assess_year} projection and alternative catch-only projection in which the 2026 assumed catches are set to 67% of a new ACL based on 1) a category 1 sigma of 0.5 and 2) actual and assumed catches through 2025 rather than fixed at the values calculated by the GMT based on the current harvest specifications. The listed buffer in 2026 is for a category 1 assessment, and does not apply to the 2023 ABC. See caption of table above for additional details.")

```

This document details a catch-only projection for `r species`. The most recent assessment for `r species` was conducted in `r assess_year`. This analysis updates catches between `r assess_year` - `r current_year - 1` to the best information about removals as provided by the Groundfish Management Team (GMT). The removals for `r current_year` and `r current_year + 1` were set equal to mortality projections provided by the GMT for each fleet in the model. For years `r current_year + 2` and beyond, removals were set equal to the projected Annual Catch Limit (ACL) based on a category `r category` time-varying $\sigma$ of `r sigma` and a P* value of `r p_star`. These buffers are the same as in the previous stock assessment, with two additional years of buffers for years `r current_year + 10` and `r current_year + 11`. 

Two alternative projections are provided. In @tbl-proj2, first, a new ACL for 2026 is calculated based on assumed catches through 2025. Then, a full projection is done using the same assumption of 67% ACL attainment and similar apportioning by fleet in 2026 as the projection in @tbl-proj, but applied to the newly calculated 2026 ACL. In @tbl-proj3, first, a new ACL for 2026 is calculated based on assumed catches through 2025 *and* a category 1 sigma of 0.5. Then, the same full projection is done assuming 67% attainment. 

```{r}
#| label: calc-table
#| echo: false
#| warning: false
#| message: false

# define ranges of years to get values
all_years <- assess_year:final_proj_year
proj_years <- (current_year + 2):final_proj_year
proj_years2 <- (current_year + 1):final_proj_year
catch_years <- assess_year:(current_year + 1)
catch_years2 <- assess_year:(current_year)
diff_years <- current_year - assess_year

# function to get model outputs
catch_only_projection_table <- function(model, all_years, proj_years, catch_years) {
    # get realized catch from model output
    catch <- model$derived_quants |>
        dplyr::filter(Label %in% paste0("ForeCatch_", catch_years)) |>
        dplyr::reframe(
            Year = stringr::str_extract(Label, "\\d+") |> as.numeric(),
            Catch = Value
        )
    ofl <- model$derived_quants |>
        dplyr::filter(Label %in% paste0("OFLCatch_", proj_years)) |>
        dplyr::reframe(
            Year = stringr::str_extract(Label, "\\d+") |> as.numeric(),
            OFL = Value
        )
    acl <- model$derived_quants |>
        dplyr::filter(Label %in% paste0("ForeCatch_", proj_years)) |>
        dplyr::reframe(
            Year = stringr::str_extract(Label, "\\d+") |> as.numeric(),
            ACL = Value
        )
    sb <- model$derived_quants |>
        dplyr::filter(Label %in% paste0("SSB_", all_years)) |>
        dplyr::reframe(
            Year = stringr::str_extract(Label, "\\d+") |> as.numeric(),
            SB = Value
        )
    status <- model$derived_quants |>
        dplyr::filter(Label %in% paste0("Bratio_", all_years)) |>
        dplyr::reframe(
            Year = stringr::str_extract(Label, "\\d+") |> as.numeric(),
            status = round(Value, 3)
        )

    # get buffer from PEPtools function
    Buffer <- PEPtools::get_buffer(
        years = all_years, sigma = sigma, pstar = p_star,
        verbose = FALSE
    ) |>
        dplyr::rename(Year = year, Buffer = buffer) |>
        dplyr::filter(Year %in% proj_years)

    # combine all the tables above (joined by year)
    table <- purrr::reduce(
        list(ofl, Buffer, acl, sb, status, catch),
        ~ dplyr::full_join(.x, .y, by = "Year")
    ) |>
        dplyr::arrange(Year) |>
        dplyr::relocate(Catch, .after = Year) |>
        dplyr::mutate(
            Buffer_from_ratio = dplyr::case_when(
                status >= model$btarg ~ round(ACL / OFL, 3),
                .default = NA # ratio doesn't work for years with 40-10 adjustment
            ),
            .after = Buffer
        ) |>
        dplyr::mutate(
            ABC = Buffer * OFL,
            .after = Buffer_from_ratio
        ) |>
        dplyr::rename(
            `Actual & Assumed Removals` = Catch,
            `Stock Status` = status,
            `Spawning Output` = SB,
        )
}

combine_tables <- function(table1, table2) {
    dplyr::full_join(
        table1,
        table2,
        by = "Year"
    ) |>
        dplyr::select(
            c(
                "Year",
                "OFL (yyyy)", "OFL",
                "Buffer",
                "ABC (yyyy)", "ABC",
                "Actual & Assumed Removals (yyyy)", "Actual & Assumed Removals",
                "ACL (yyyy)", "ACL",
                "Spawning Output", "Stock Status (yyyy)", "Stock Status"
            )
        ) |>
        dplyr::rename_with(.fn = function(x) {
            gsub("yyyy", assess_year, x)
        })
}

# function to apply gt formatting to the table
format_table <- function(table) {
    table |>
        gt::gt() |>
        gt::fmt_number(
            columns = tidyselect::starts_with("Actual") | tidyselect::starts_with("OFL") | tidyselect::starts_with("ABC") | tidyselect::starts_with("ACL"),
            decimals = 0
        ) |>
        gt::fmt_number(
            columns = tidyselect::contains("Spawning"),
            decimals = spawn_output_decimals
        ) |>
        gt::tab_options(
            table.font.size = 11, # reduced font size
            latex.use_longtable = TRUE
        ) |>
        gt::sub_missing(
            columns = tidyselect::everything(),
            missing_text = "---"
        ) |>
        gt::cols_align(
            align = "center"
        ) |>
        gt::cols_width(
            everything() ~ px(65)
        ) |>
        gt::data_color(
            columns = tidyselect::contains(paste(assess_year)),
            palette = "gray90",
            na_color = "gray90"
        ) |>
        gt::as_latex()
}

table <- catch_only_projection_table(model,
    all_years = all_years, proj_years = proj_years, catch_years = catch_years
)
table_original <- catch_only_projection_table(model_original,
    all_years = head(all_years, 12), proj_years = proj_years - diff_years, catch_years = catch_years - diff_years
)
table2 <- catch_only_projection_table(model2,
    all_years = all_years, proj_years = proj_years2, catch_years = catch_years2
)
table3 <- catch_only_projection_table(model3,
    all_years = all_years, proj_years = proj_years2, catch_years = catch_years2
)
```

```{r}
#| label: warning-about-buffer
#| echo: false
#| warning: true
#| message: true

if (any(abs(table$Buffer_from_ratio - table$Buffer) > 0.01, na.rm = TRUE)) {
    cli::cli_alert_warning("ACL / OFL differs from buffer calculated by PEPtools::get_buffer() by more than 0.01 for years with status > target. See values below.")
    print(table)
} else {
    table <- dplyr::select(table, -Buffer_from_ratio)
}

if (any(abs(table_original$Buffer_from_ratio - table_original$Buffer) > 0.01, na.rm = TRUE)) {
    cli::cli_alert_warning("ACL / OFL differs from buffer calculated by PEPtools::get_buffer() by more than 0.01 for years with status > target. See values below.")
    print(table_original)
} else {
    table_original <- dplyr::select(table_original, -"Buffer_from_ratio")
}

table_original <- table_original |>
    dplyr::select(-Buffer, -"Spawning Output") |>
    dplyr::rename_with(~ paste(.x, "(yyyy)"), .cols = -Year)
# combine the two tables
wide_table <- combine_tables(table, table_original)

# alternative COP 1
wide_table2 <- combine_tables(table2, table_original)
# move 2026 ACL to assumed catch column
wide_table2$"Actual & Assumed Removals"[wide_table2$Year == 2026] <- 
  wide_table2$ACL[wide_table2$Year == 2026]
# calculate 2026 ACL from ABC
# get 2026 stock status for use in 40-10 or 25-5 rule
D <- wide_table2$"Stock Status"[wide_table2$Year == 2026]
#
adj <- 1.0
# 40-10 rule for most species
if (D < model$btarg & model$btarg == 0.40) {
  adj <- (4/3)*(D-0.1)/D
}
# 25-5 rule for flatfish
if (D < model$btarg & model$btarg == 0.25) {
  adj <- (25/20)*(D-0.05)/D
}
# calculate ACL from adjustment * ABC
wide_table2$ACL[wide_table2$Year == 2026] <- 
  adj * wide_table2$ABC[wide_table2$Year == 2026]

# alternative COP 2
wide_table3 <- combine_tables(table3, table_original)
# move 2026 ACL to assumed catch column
wide_table3$"Actual & Assumed Removals"[wide_table3$Year == 2026] <- 
  wide_table3$ACL[wide_table3$Year == 2026]

# calculated 2026 ABC from OFL
wide_table3$ABC[wide_table3$Year == 2026] <- 
  wide_table3$OFL[wide_table3$Year == 2026] * 
  PEPtools::get_buffer(2023:2026, sigma = 0.5, pstar = 0.45, verbose = FALSE)[4, 'buffer']

# calculate 2026 ACL from ABC
# get 2026 stock status for use in 40-10 or 25-5 rule
D <- wide_table3$"Stock Status"[wide_table3$Year == 2026]
#
adj <- 1.0
# 40-10 rule for most species
if (D < model$btarg & model$btarg == 0.40) {
  adj <- (4/3)*(D-0.1)/D
}
# 25-5 rule for flatfish
if (D < model$btarg & model$btarg == 0.25) {
  adj <- (25/20)*(D-0.05)/D
}
# calculate ACL from adjustment * OFL
wide_table3$ACL[wide_table3$Year == 2026] <- 
  adj * wide_table3$ABC[wide_table3$Year == 2026] #ABC

wide_table3$Buffer[wide_table3$Year == 2026] <- PEPtools::get_buffer(2023:2026, sigma = 0.5, pstar = 0.45, verbose = FALSE)[4, 'buffer']
```

::: {.landscape}

```{r}
#| label: tbl-proj
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| tbl-cap: !expr table_caption
format_table(wide_table)
```

{{< pagebreak >}}

```{r}
#| label: tbl-proj2
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| tbl-cap: !expr table_caption2
format_table(wide_table2)
```

{{< pagebreak >}}

```{r}
#| label: tbl-proj3
#| echo: false
#| warning: false
#| message: false
#| eval: true
#| tbl-cap: !expr table_caption3
format_table(wide_table3)
```
