```{r executive, echo = FALSE}
library(here)
library(dplyr)
library(tidyr)

executive <- list()
executive[["stock"]] <- paste0("This assessment reports the status of ",
  spp, " (_", spp.sci, "_) off the ", coast, 
  " coast using data through xxxx.")

tab_dir1 <- here::here("doc", "FinalTables", "Summary")

```

# Executive summary{-}

## Stock{-}

This assessment applies to Shortspine thornyhead (Sebastolobus alascanus) off of the west coast of the United States from the U.S.-Canada border to the U.S.-Mexico border using data through 2023. Shortspine thornyheads are modeled in this assessment as a single stock. They have been reported as deep as 1,524 m, and this assessment applies to their full depth range although survey and fishery data are only available down to 1,280 m. 

## Catches{-}

Landings of shortspine are estimated to have risen to a peak of 4,815 mt in 1989, followed by a sharp decline during a period of trip limits and other management measures imposed in the 1990s. Since the institution of separate trip limits for shortspine and longspine thornyheads, the fishery had more moderate removals of between 1,000 and 2,000 mt per year from 1995 through 1998. Landings fell below 1,000 mt per year from 1999 through 2006, then rose to 1,531 in 2009 and have declined since that time. Recreational fishery landings of thornyheads were negligible, so only commercial landings were included in the model. 

Commercial landings are dominantly bottom trawl caught, and non-trawl landings include all other gear types, the majority of which is longline. For this assessment, trawl landings were divided into North (the waters off Washington and Oregon) and South (the waters off California) fleets, and non-trawl landings represent one, coastwide fleet. Discard rates (landings divided by total catch) for shortspine have been estimated as high as 43% per year, but are more frequently below 20%. Discard rates in the trawl fleets, available from the West Coast Groundfish Observer Program (WCGOP), declined from 2003–2011 and have since increased from 2012-present, when the catch shares system began. 

```{r, results='asis'}

catch_tab<-read.csv(file.path(tab_dir1, "a_Catches_ES.csv"))
names(catch_tab)<-c("Year", "North Trawl", "South Trawl", "Non-Trawl", "Total Landings", "Total Dead")  
sa4ss::table_format(x = catch_tab,
                  caption = "Recent landings by fleet, total landings summed across fleets, and the total mortality including discards.",
                  label = "catches")
```

```{r, results = "asis"}
# Figure 3 Catch history
sa4ss::add_figure(filein = file.path(fig_dir3, "catch2 landings stacked.png"), 
                  caption = "Estimated landing history for shortspine thornyhead.", 
                  label = "catch_hist")
```


## Data and assessment{-}

The most recent assessment for shortspine thornyhead was conducted in 2012 [@taylor_stephens_2013]. Stock status was determined to be above the target biomass and catches did not attain the full management limits so reassessment of thornyheads has not been a higher priority. This assessment used Stock Synthesis [@methot_stock_2013] Version 3.30.21, used in many other recent west coast assessments. 
					
Data were divided into three fishery fleets: North trawl (the waters off Washington and Oregon), South trawl (the waters off California), and coastwide non-trawl, and three survey fleets: the Alaska Fishery Science Center (AFSC) Triennial Shelf Survey from 1980-2003, which was divided into early (pre-1995) and late period (post-1995) to account for a change in depth-sampling, and the West Coast Groundfish Bottom Trawl Survey (WCGBTS), 2004-2022. 

Most data used in the 2012 assessment were newly pulled and processed for this assessment, including length compositions from all fishing and survey fleets, indices of abundance derived from new geostatistical analyses, discard rates from both a 1980s observer study (Pikitch et al., 1988) and the current WCGOP, historical catch data from Washington, Oregon, and California, and all reported catches from 1981-2022. The only data  taken from the previous assessment without reanalysis were discard rates from the Enhanced Data Collection Project (EDCP) study in the 1990s. 

New maturity analyses of samples collected in the WCGBTS in 2011, 2013, 2014, 2016 and 2018 were available for this assessment (M. Head, pers. comm.). The larger number and better spatial coverage of these samples allowed the use of statistical modeling to better understand the spatial variation in the proportion of female spawning. This assessment also assumes a new fecundity relationship, in which fecundity increases with body size. New growth curves were estimated, using data from @butler_1995, which were similar to the curves assumed in the 2005 and 2013 assessments. In the previous assessment, a Beverton-Holt stock recruitment relationship was assumed and steepness was fixed at 0.60. This assessment fixed steepness at 0.72, as recommended by @thorson_steepness_2019. Natural mortality was also slightly updated from the 2013 assessment to be fixed at 0.04. 

This assessment estimated 180 parameters. The log of the unfished equilibrium recruitment, log(R0), controls the scale of the population and annual deviations around the stock-recruit curve (135 parameters) allow for more uncertainty in the population trajectory. In addition, 43 selectivity and retention parameters for the three fishery fleets and three surveys allowed for estimation of annual length compositions and discards rates. Two catchability parameters were analytically computed from the data, and one additional parameter, representing additional variability in the early Triennial survey, was directly estimated by the model.

## Stock biomass and dynamics{-}

Unfished equilibrium spawning output (B0) is estimated to be 20.332 trillion eggs, with a 95% confidence interval of 16.338-24.327 trillion eggs. The B0 estimate here is not comparable to previous assessment as the integration of new fecundity and maturity assumptions have changed the output units from traditional biomass to spawned  eggs. Spawning biomass is estimated to have remained stable until the early-1970s before beginning to decline near linearly through the present day. The estimated spawning output in 2023 is 8.372 trillion eggs, which represents a stock status or “depletion” (represented as spawning biomass in 2023, B2023, divided by B0) of 41.4% (Figure XX). The depletion in 2013 was estimated to be 43.6%, a large decrease from what was estimated in 2013 (~75%).

```{r, results='asis'}
ssb_tab<-read.csv(file.path(tab_dir1, "b_SSB_ES.csv"))
names(ssb_tab)<-c("Year", "Spawning Output", "Spawning Output 95% CI", "Fraction Unfished", "Fraction Unfished 95% CI")                 
# 
sa4ss::table_format(x = ssb_tab,
                  caption = "Spawning output (millions of eggs) and fraction unfished with associated 95\\% confidence intervals (CI) from the base model.",
                   label = "ssb")
```

```{r, results = "asis"}
sa4ss::add_figure(filein = file.path(fig_dir3, "ts7_Spawning_output_with_95_asymptotic_intervals_intervals.png"), 
                  caption = "Estimated spawning output trajectory for shortspine thornyhead.", 
                  label = "ssb_trajectory")
```

## Recruitment{-}

This assessment assumed a Beverton-Holt stock recruitment relationship. Steepness (the fraction of expected equilibrium recruitment associated with 20% of equilibrium spawning biomass) was fixed at 0.7, slightly higher than what was assumed in previous assessments (h=0.60). The scale of the population is largely determined by the log of unfished recruitment (R0), which was estimated to be 9.354. This results in an unfished recruitment of 11,550,000 recruits (9,281,000–13,820,000). Recruitment deviations were estimated for the years 1901 through 2022, and ranged from -0.5 to 1.5 on the log scale. Estimated recruitments do not show high variability, and the uncertainty in each estimate is greater than the variability between estimates.

```{r, results='asis'}
rec_tab<-read.csv(file.path(tab_dir1, "c_Recr_ES.csv"))
names(rec_tab)<-c("Year", "Recruitment", "95% CI", "RecDevs", "RecDev 95% CI")    
sa4ss::table_format(x = rec_tab,
                  caption = "Estimated recent trend in recruitment and recruitment deviations and the 95\\% confidence intervals (CI) from the base model.",
                  label = "rec")
```

```{r, results = "asis"}
sa4ss::add_figure(filein = file.path(fig_dir3, "ts11_Age-0_recruits_(1000s)_with_95_asymptotic_intervals.png"), 
                  caption = "Estimated recruitment timeseries.", 
                  label = "rec_trajectory")
```

## Exploitation status{-}

The summary harvest rate (total catch divided by age-1 and older biomass) closely follows the landings trajectory. The harvest rates are estimated to have never exceeded 5% and have remained below 2% in the past decade. Expressing exploitation rates in terms of spawning potential ratio (SPR) indicates that the exploitation consistently exceeded the SPR50% reference point from 1980-2018. However, the stock status is estimated to have never fallen below the B40% management target, though the uncertainty interval around the 2023 estimate does encapsulate the B40% target.

```{r, results='asis'}
spr_tab<-read.csv(file.path(tab_dir1, "d_SPR_ES.csv"))
names(spr_tab)<-c("Year", "(1-SPR)/(1-SPR 50%)", "95% CI", "Exploitation Rate", "95% CI")    
sa4ss::table_format(x = spr_tab,
                  caption = "Estimated recent trend in the (1-SPR)/(1-SPR 50\\%) where SPR is the spawning potential ratio the exploitation rate, and the  95\\% intervals.",
                  label = "spr")
```

```{r, results = "asis"}
sa4ss::add_figure(filein = file.path(fig_dir3, "ts9_Relative_spawning_output_intervals.png"), 
                  caption = "Estimated relative spawning output trajectory for shortspine thornyhead.", 
                  label = "rel_ssb_trajectory")

sa4ss::add_figure(filein = file.path(fig_dir3, "ts_summaryF.png"), 
                  caption = "Summary F rate.", 
                  label = "summary_f")

sa4ss::add_figure(filein = file.path(fig_dir3, "SPR3_ratiointerval.png"), 
                  caption = "Estimated spawning potential ratio.", 
                  label = "spr_trajectory")

sa4ss::add_figure(filein = file.path(fig_dir3, "SPR4_phase.png"), 
                  caption = "Phase diagram.", 
                  label = "phase_diagram")

```

## Ecosystem considerations{-}

Replace text with
a summary of reviewed environmental and ecosystem factors that appear to be correlated with stock dynamics.
These may include variability in they physical environment, habitat, competitors, prey, or predators that directly or indirectly affects the stock's status, vital rates (growth, survival, productivity/recruitment) or range and distribution.
Note which, if any, ecosystem factors are used in the assessment and how (e.g., as background information, in data preparations, as data inputs, in decisions about model structure).

## Reference points{-}

Replace text with
management targets and definition of overfishing, including the harvest rate that brings the stock to equilibrium at $B_{40\%}$, i.e., the $B_{MSY}$ proxy and the equilibrium stock size that results from fishing at the default harvest rate, i.e., the $F_{MSY}$ proxy.
Include Table of estimated reference points for ssb, SPR, exploitation rate, and yield based on SSB proxy for MSY, SPR proxy for MSY, and estimated MSY values.

## Management performance{-}

Include Table of most recent 10 years of
catches in comparison with OFL, ABC, HG, and OY/ACL values,
overfishing levels,
actual catch and discard.
Include OFL (encountered), OFL (retained), and OFL (dead) if different due to discard and discard mortality.

## Unresolved problems and major uncertainties{-}

Major uncertainties in the model are centered around uncertainty in biological processes including growth, maturity, and mortality. The absence of reliable ageing methods for Shortspine thornyhead, particularly, makes it difficult to estimate growth and natural mortality. Sensitivities demonstrated that changes to the growth curve have large effects on the estimated stock status. Likelihood profiles over natural mortality demonstrate the model to be quite sensitive to its assumed value. There is insufficient information in the data to estimate mortality directly, constraining us to use meta-analyses or other mortality estimators, which frequently make use of aging information that is unavailable, and again, highly uncertain for shortspine thornyhead. Due to imperfect seasonal and spatial coverage of histological data for shortspine thornyhead, there is significant uncertainty about the shape of the species’ maturity curve, though the model appears to be largely insensitive to variations in maturity.

This model fails to fully capture the observed increase in abundance seen in the WCGBTS index time series in 2021 and 2022. The model also fails to fully capture the peak of the length compositions for the Northern Trawl fleet, underestimating the number of mid-sized fish that the fleet takes *(Figure XX)*. This underestimation appears to be consistent, particularly in the last 10 years *(Figure XX)*, implying a possible recent change in selectivity.


## Decision table and projections{-}

Replace text with
projected yields (OFL, ABC, and ACL), spawning biomass, and stock depletion levels for each year.
OFL calculations should be based on the assumption that future catches equal ABCs and not OFLs.

## Scientific uncertainty{-}

Replace text with
the sigma value and the basis for its calculation.

## Research and data needs{-}

Research and data needs for future assessments include the following: 

1. Research into ageing methods and availability of reliable age data would be valuable for future stock assessments. Otoliths have been collected in good quantities from the NWFSC survey, but there is currently no validated ageing method for Shortspine thornyhead. 
2. More investigation into maturity of Shortspine thornyhead is necessary to understand the patterns in maturity observed in WCGBTS samples. 
3. Information on possible migration of Shortspine thornyheads would be valuable for understanding stock dynamics. Analysis of trace elements and stable isotopes in shortspine otoliths may provide valuable information on the extent of potential migrations. Possible connections between migration and maturity could likewise be explored. 
4. A greater understanding of the connection between thornyheads and bottom type could be used to refine the indices of abundance. Thornyheads are very well sampled in trawlable habitat, but the extrapolation of density to a survey stratum could be improved by accounting for the proportion of different bottom types within a stratum and the relative density of thornyheads within each bottom type.
5. Additional investigation into spatial stock structure could be valuable for determining whether future assessments should develop a spatial assessment model, or if shortspine thornyhead should be assessed at distinct spatial scales in the future.

